Estimation of the effect of a continuous treatment variable

### 2025-10-28 Writing/manuscript_CF_continuous.qmd
Draft of chapter two

---
title: "Estimation of the effect of a continuous treatment variable"
format:
  docx:
    toc: true

execute:
  echo: false
bibliography: references.bib
---

@@

We compare six estimators—CF, SO_ANN, DO_ANN, NO_ANN, SO_RF, and NO_RF—over 500 Monte-Carlo replications and five training sizes (1, 3, 5, 10 and 20 fields). Each replication contains 1,440 evaluation units. Accuracy is assessed by the RMSE of the estimated economically optimal nitrogen rate (EONR) with respect to the true (simulated) EONR. Figures 1–4 summarize errors; Figure 5 panels illustrate representative yield–N response functions that help interpret the error patterns.

@@

@fig-rmse-hist shows a clear ordering across methods. CF concentrates tightly at low RMSE for all training sizes, indicating both accuracy and stability. SO_ANN shifts steadily left as the number of fields increases, with visibly narrower spreads at 5 and 10 fields. In contrast, NO_ANN exhibits a heavy right tail with a single field but collapses toward the leading methods at 5–10 fields, showing its sensitivity to sample size. SO_RF improves with more fields yet remains less accurate than SO_ANN or CF. NO_RF has broad, right-shifted distributions and is a weak baseline. DO_ANN performs poorly across all sizes, with distributions centered at very high RMSE and little improvement from additional fields.

@@

Mean RMSE by model and training size (@fig-avg-performance) confirms these patterns. CF is consistently low (≈15 across sizes). SO_ANN improves monotonically as data increase and attains the best mean at 10 fields (≈11). NO_ANN is unusable with a single field (mean ≈82) but becomes competitive by 5–10 fields (≈14 and ≈13). SO_RF yields moderate means (≈22 at 10 fields), while NO_RF remains highest among the non-ANN baselines (≈36–40). DO_ANN is uniformly poor (≈90–93). The heatmap (Fig. 3) prints the means in each cell and emphasizes the strong gains from orthogonalization (SO_ANN and CF) relative to their non-orthogonal counterparts.

@@

Using NO_RF as a baseline, CF reduces mean RMSE by roughly 58–61 % across training sizes (@fig-improvement). SO_ANN delivers comparable or larger gains (≈54–69 %), with the largest improvement at 10 fields. SO_RF provides modest gains (≈38–42 %). NO_ANN is worse than the baseline with 1 field but improves substantially at 5–10 fields (≈63–65 % reduction). DO_ANN is consistently inferior to the baseline at all sizes.

@@

```
```{r}

@@

best_models %>%
  gt() %>%
  tab_header(
    title = "Best models by #field",
    subtitle = "Ordered by lowest mean RMSE"
  ) %>%
  fmt_number(columns = c(mean), decimals = 3) %>%
  cols_label(
    field_f = "# fields",
    mean = "mean RMSE"
  ) %>%
  tab_options(table.font.size = "small")

@@

Predicted yield–N curves (@fig-yield-res-1 through @fig-yield-res-3) for five randomly selected analysis units clarify why methods differ. SO_ANN produces agronomically plausible, monotone-concave responses with clear plateaus (ten-field panel), yielding interior optima and low EONR error. NO_ANN resembles SO_ANN once trained on 5–10 fields, but with a single field it is noticeably noisier, aligning with its heavy-tailed errors. Tree-based models (NO_RF and SO_RF) generate step-like, piecewise-constant responses; the implied EONR can only move at jump points, which coarsens the optimum and limits accuracy. DO_ANN often yields nearly linear increasing curves across the tested N range, pushing the optimum to the upper bound and explaining its uniformly high error. Although CF is not naturally expressed as per-unit yield curves, its error distribution indicates high stability across all sizes.

@@

@@

@@

Across 500 replications, SO_ANN (with ≥10 fields) and CF deliver the best overall accuracy. Orthogonalization is critical: SO_ANN and CF are consistently superior to NO_ANN and NO_RF, especially in small-to-moderate samples. NO_ANN becomes competitive only once training data reach at least five fields. Tree-based baselines remain substantially less accurate, and the doubly-orthogonal ANN (DO_ANN) is inappropriate for this setting because it fails to recover the saturating response shape. These findings are robust across training sizes and are consistent.

>>>>

---
title: "Estimation of the effect of a continuous treatment variable"
format:
  docx:
    toc: true
    number-sections: true
    number-tables: true 
  html:
    toc: true
execute:
  echo: false
bibliography: references.bib
---

@@

We compare six estimators—CF, SO_ANN, DO_ANN, NO_ANN, SO_RF, and NO_RF—over 500 Monte-Carlo replications and five training sizes (1, 3, 5, 10 and 20 fields). Each replication contains 1,440 evaluation units. Accuracy is assessed by the RMSE of the estimated economically optimal nitrogen rate (EONR) with respect to the true (simulated) EONR.

@@

Taken together, @fig-rmse-hist and @fig-avg-performance illustrate clear performance patterns across methods and training sizes. CF delivers low and tightly concentrated RMSE distributions even with limited data—consistently maintaining strong accuracy and stability for 1–5 fields (mean RMSE ≈ 14.7, 13.1, and 12.1, respectively), with only modest additional gains at 10 and 20 fields (11.9–11.8). SO_ANN initially lags behind CF (18.4 at one field), but its mean RMSE decreases steadily as more fields are added, surpassing CF performance at 10 and 20 fields (11.2 and 8.5, respectively). This pattern underscores the neural network’s superior ability to exploit richer spatial and agronomic heterogeneity when sufficient training data are available.

By contrast, NO_ANN exhibits substantial instability in the single-field case, with a heavy right tail and much higher errors, but rapidly converges toward the leading methods by 5–10 fields (mean RMSE improving to around 14–13), highlighting its strong dependence on sample size. SO_RF improves modestly with more fields (mid-20s RMSE for 10–20 fields) but remains consistently less accurate than SO_ANN or CF. NO_RF maintains broad, right-shifted distributions and higher mean errors throughout (≈ 36–40), confirming its role as a weak baseline. Finally, DO_ANN performs poorly throughout, with both histograms and means showing persistently high errors (≈ 90+) and negligible benefit from additional data, reflecting over-correction in the orthogonalization process.

@@

@@

@fig-improvement reports each method’s percentage improvement in mean RMSE relative to the NO_RF baseline. CF consistently provides the strongest gains among the robust approaches when data are scarce—reducing error by ≈63% with a single field, and climbing to ≈67–68% once 5–20 fields are available. SO_ANN exhibits a similar improvement trend but grows even more dominant as more diverse data accumulate, achieving ≈53% improvement at 1 field, ≈63% by 5 fields, and reaching ≈68% at 10 fields and ≈77% at 20 fields, the highest improvement across all models. NO_ANN transitions from a large negative result when trained on just one field (substantially worse than baseline) to strong relative improvements exceeding 65% by 10–20 fields, reinforcing its reliance on adequate sample support. SO_RF offers moderate improvement (≈38–43%) that changes little with added fields. DO_ANN remains entirely uncompetitive across all sample sizes—showing worsening performance relative to NO_RF by more than 100–150%, indicating severe model misspecification.

Together, these results emphasize that while CF is the most reliable option under limited data availability, SO_ANN ultimately yields the largest relative error reduction when multi-field datasets are accessible, highlighting its superior scalability with richer training information.

@@

```
@tbl-best_model_by_field_count highlights the top-performing model for each training size. CF is the dominant method when only a few fields are available, delivering the lowest mean RMSE for 1–5 fields. As field diversity increases, SO_ANN begins to outperform CF, becoming the preferred approach at 10 fields and showing a substantial further accuracy gain at 20 fields.

@@

best_models %>%
  gt() %>%
  tab_header(
    title = "Best models by field count",
    subtitle = "Ordered by lowest mean RMSE"
  ) %>%
  fmt_number(columns = c(mean), decimals = 3) %>%
  cols_label(
    field_f = "# fields",
    mean = "mean RMSE"
  ) %>%
  tab_options(table.font.size = "small")

@@

@@

@fig-yield-res-combined displays predicted and true yield response functions for the best-performing models at each training size. For the CF, shown for 1, 3, and 5 fields, the estimated curves closely track the overall shape of the true response but exhibit a slight upward bias in predicted yield levels. The general curvature, including diminishing returns at higher nitrogen rates, is captured consistently even under data-constrained conditions. As more fields are incorporated, the CF predictions become smoother and more aligned with the true function, reflecting improved generalization.

For the SO_ANN, shown for 10 and 20 fields, the estimated response matches the true curve more closely in both slope and curvature, particularly at higher nitrogen levels where CF tended to flatten. At 20 fields, SO_ANN produces the most accurate approximation overall, with prediction and truth nearly coincident across the nitrogen range. These differences visually reinforce the earlier quantitative results: CF offers strong robustness with sparse multi-field data, while SO_ANN increasingly excels as richer spatial heterogeneity becomes available, enabling superior recovery of the underlying agronomic yield response function.

@@

# Discussion

This study evaluates alternative causal-ML strategies for estimating heterogeneous yield response to a continuous nitrogen treatment in an OFPE setting. CF consistently attains low and tightly concentrated EONR errors with only 1–5 training fields, indicating strong small-sample stability (see @fig-rmse-hist and @fig-avg-performance). SO_ANN improves monotonically with additional fields and ultimately dominates once ≥10 fields are available, achieving the lowest mean RMSE overall (e.g., 11.2 at 10 fields and 8.5 at 20 fields). These complementary strengths that CF is preferred under data-constrained conditions, whereas SO_ANN becomes the method of choice as multi-field datasets accumulate (see @tbl-best_model_by_field_count).

Methodologically, the contrast between single-orthogonalization and double-orthogonalization is central. For continuous treatments, the double-orthogonal residualization of the treatment does not generally preserve the relevant heterogeneous response structure. Our simulations reflect this: DO_ANN performs poorly and does not benefit from larger samples. In contrast, single-orthogonalization—residualizing only the outcome to remove baseline variation from covariates $W$—retains the full interaction between $X$ and $N$, allowing the final learner to recover the nonlinearity and heterogeneity of the response function. The extended CF strikes a different balance: by expressing $N$ through a spline basis and orthogonalizing the basis components, it restores a linear-in-parameters score that CF can learn reliably.

@@

By combining orthogonalization with flexible final-stage learners, we demonstrate how to preserve the strengths of causal inference while accommodating continuous treatments and agronomic heterogeneity.

Across 500 simulated test fields, two approaches consistently dominate: the extended causal forest and the single-orthogonalized neural network. CF is highly robust in small-sample environments, achieving strong accuracy even with training data from just one to five fields. As the number of training fields increases, SO_ANN increasingly improves as a result of spatial and agronomic diversity, ultimately delivering the best predictive accuracy of all methods once ten or more fields are available. In contrast, non-orthogonalized baselines require larger samples to become competitive and the doubly-orthogonalized ANN performs poorly in all conditions due to the incompatibility between double-orthogonalization and continuous treatments. Methodologically, our results highlight the importance of aligning the orthogonalization strategy with the structure of the treatment variable in causal ML applications.
